\scuba: iterative map-making with the Sub-Millimetre User Reduction Facility

### 2012-11-30 smurf_paper.tex
update text and plots for point source with extinction correction fix

The accurate measurement of positions and brightnesses of known point
sources are necessary in real-time to establish telescope pointing
offsets and focus. They are also necessary to measure the FCF
(absolute calibration), and hence noise performance of the instrument
in astronomically-useful units.  In this example we produce a
450\,\micron\ map of Uranus (observation 26 on 20111017), which is a
nearly point-like source for \scuba\ that is commonly used as a
primary flux calibrator. The CV daisy pattern was used, with a scan
speed of 155 arcsec\,sec$^{-1}$. We perform several different
reductions of the data to illustrate the purpose of various model
components and the convergence properties of the solution
(Fig.~\ref{fig:pointmaps}). In all cases the maps are produced on a
grid of azimuth (horizontal) and elevation (vertical) offsets from the
position of Uranus (the origin), using 2\,arcsec pixels.

The first, simplest reduction of the data uses only the \model{COM}
model to estimate and remove the common-mode signal in order to
suppress low-frequency noise in the data. After \model{COM}, the
extinction correction is applied (\model{EXT}), and an initial map is
estimated using equal weighting for all of the detectors. This
estimate of \model{AST} is then removed from the data, and the noise
is measured in the residuals to estimate weights for the subsequent
and final iteration. The resulting map after these two iterations is
shown in Fig.~\ref{fig:pointmaps}a. While the \snr\ of Uranus is
clearly large ($\sim$250), enabling us to see the faint sidelobes
(circle and cross pattern), the map also has obvious circular streaks
and other large-scale ripples. The circular streaks are due to the
fact that \model{COM} does not account for all of the low-frequency
(i.e., uncorrelated) noise (see Fig.~\ref{fig:pspec}), and therefore
each bolometer leaves a trace of the circular scan pattern in the map,
as their baselines slowly drift independently. A significant
contribution to the larger-scale ripples in the map, however, can be
made by degeneracies in the map solution, as we discuss below.

To illustrate how large-scale ripples can form (and grow), the same
map solution is run for 100 iterations and shown in
Fig.~\ref{fig:pointmaps}b, now exhibiting a strong vertical
gradient. The degeneracy is easy to understand if the time-domain
behaviour of each model component is considered. The top panel of
Fig.~\ref{fig:degeneracy} shows the residual signals for a single
bolometer after 2 (black) and 100 (grey) iterations, which are nearly
identical, yet the change in the estimated \model{COM} (green) and
\model{AST} (red) signals between 2 and 100 iterations are
large. However, it is also clear that the estimated \model{COM} and
\model{AST} signals are complimentary. In other words, the large
change in the \model{AST} signal is cancelled by freedom in the
\model{COM} signal to grow with opposite sign. For comparison, the
bottom panel of Fig.~\ref{fig:degeneracy} shows the telescope pointing
for this section of data, and the shapes of the \model{AST} and
\model{COM} signals match the elevation component, which is aligned
with the gradient in Fig.~\ref{fig:pointmaps}b. Generically, the
calculation of \model{COM} will remove any information on angular
scales that are larger than the array footprint (outline shown in
Fig.~\ref{fig:pointmaps}b for reference), meaning that the map
solution is unconstrained on such large scales.

Attacking the problem of streaks, a simple method of removing residual
(un-correlated) sources of low-frequency noise is to apply a high-pass
filter after the common-mode removal. A third reduction of the data
uses the baseline map-making parameters, as described in
Section~\ref{sec:algorithm}, which adds the \model{FLT} model to
accomplish this task immediately prior to map estimation. We set the
filter edge based on an angular scale of 200\,arcsec, which, given the
scan speed of 155\,arcsec\,sec$^{-1}$, corresponds to a frequency of
0.78\,Hz (note in Fig.~\ref{fig:pspec} that this frequency is slightly
above the $1/f$ knee at 450\,\micron\ after common-mode removal). Now
using the automatic map-based convergence test
(Section~\ref{sec:converge}), the solution converges after 10
iterations (Fig.~\ref{fig:pointmaps}c). Both the circular streaks and
the large-scale gradient are now removed, but they have been replaced
by an obvious, circularly-symmetric ringing pattern about Uranus. The
reason for this ringing is that the hard-edged high-pass filter in
frequency space is equivalent to a $\sinc$ function-like response in
map space. Since the scan pattern is fairly isotropic (scans at all
position angles), and there is a bright point-like source at the
centre, the result is an azimuthally-symmetric $\sinc$ function-like
pattern in the map. The gradient resulting from the degeneracy between
the common-mode and map has been removed because the filter scale of
200\,arcsec (3.3\,arcmin) is smaller than the array field-of-view
($\sim$5\,arcmin). If it were greater, degeneracies would again begin
to appear on those larger scales.

This example illustrates the need for constraints in the map solution
in many situations. For calibrators (and other previously known
bright, compact sources), a good, simple prior is to constrain the map
to a value of zero away from the known locations of emission (part of
the \model{AST} model calculation, Section~\ref{sec:ast}).  In
Fig.~\ref{fig:pointmaps}d, a solution is produced in an identical
manner to Fig.~\ref{fig:pointmaps}c, but now setting the map
explicitly to zero beyond a radius of 60\,arcsec from the location of
Uranus (much larger than the FWHM of the main lobe), for all but the
final iteration. In this case, the map converges after 6 iterations,
and the ringing has been effectively removed. Since the map is now
flat away from the source, and constrained to a value of zero, it is
appropriate for performing aperture photometry directly, with no need
for an additional reference annulus.

The reason this prior works is simple to understand if the bolometer
data are thought of as providing useful differential measurements up
to scales corresponding to the filter edge. In this example, the
bolometers retain information on brightness differences up to
200\,arcsec scales. Since the map is constrained to zero to within
60\,arcsec of the peak of Uranus (well within 200\,arcsec), the
solution is then able to accurately reconstruct the differential peak
intensity of Uranus with respect to this constrained background.  This
approach to map-making is similar to that employed for poorly
cross-linked scans of compact (though resolved) sources by
\citet{wiebe2009} using BLAST data.

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{pointmaps}
\caption{Multiple reductions of a 450\,\micron\ CV daisy scan of
  Uranus, all scaled between $-$0.002\,pW (white) and +0.002\,pW
  (black). (a) Reduction in which only common-mode subtraction is used
  to suppress low-frequency noise, and the reduction is forced to use
  2 iterations (after the first iteration an estimate of the source
  flux is removed, and the noise is measured in the residuals to
  obtain appropriate weighting for the second and final
  iteration). Circular streaks are caused by independent low-frequency
  noise that is not removed by the common-mode (Uranus peak
  0.15\,pW). (b) Same as (a) but using 100 iterations, illustrating the
  degeneracy between large-scale structure and the common-mode (the
  footprint of working bolometers is also shown for reference, Uranus
  peak 0.15\,pW). (c) Reduction in which high-pass filtering above
  0.775\,Hz (corresponding to a spatial scale of 200\,arcsec, as
  indicated) is applied after common-mode removal, but before the map
  estimate. The map-based convergence test is activated and the
  solution halts after 10 iterations, but leaving large-scale ringing
  due to the filter (Uranus peak 0.27\,pW). (d) Same as (c), but now
  the region of the map beyond the white dot-dashed circle is
  constrained to a value of zero until all but the final iteration.
  The map is now extremely flat, there is little attenuation of the
  source flux, and the diffraction pattern is clearly seen (Uranus
  peak 0.27\,pW).}
\label{fig:pointmaps}
\end{figure*}

>>>>

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{pointmaps}
\caption{Multiple reductions of a 450\,\micron\ CV daisy scan of
  Uranus, all scaled between $-$0.002\,pW (white) and +0.002\,pW
  (black). In all cases extinction correction has been applied
  immediately after common-mode removal. (a) Reduction in which only
  common-mode subtraction is used to suppress low-frequency noise, and
  the reduction is forced to use 2 iterations (after the first
  iteration an estimate of the source flux is removed, and the noise
  is measured in the residuals to obtain appropriate weighting for the
  second and final iteration). Circular streaks are caused by
  independent low-frequency noise that is not removed by the
  common-mode (Uranus peak 0.358\,pW). (b) Same as (a) but using 100
  iterations, illustrating the degeneracy between large-scale
  structure and the common-mode (the footprint of working bolometers
  is also shown for reference, Uranus peak 0.358\,pW). (c) Reduction
  in which high-pass filtering above 0.775\,Hz (corresponding to a
  spatial scale of 200\,arcsec, as indicated) is applied after
  common-mode removal, but before the map estimate. The map-based
  convergence test is activated and the solution halts after 10
  iterations, but leaving large-scale ringing due to the filter
  (Uranus peak 0.353\,pW). (d) Same as (c), but now the region of the
  map beyond the white dot-dashed circle is constrained to a value of
  zero until all but the final iteration.  The map is now extremely
  flat, there is no attenuation of the source flux compared with the
  first two reductions, and the diffraction pattern is clearly seen
  (Uranus peak 0.358\,pW).  }
\label{fig:pointmaps}
\end{figure*}

The accurate measurement of positions and brightnesses of known point
sources are necessary in real-time to establish telescope pointing
offsets and focus. They are also necessary to measure the FCF
(absolute calibration), and hence noise performance of the instrument
in astronomically-useful units.  In this example we reduce a
450\,\micron\ map of Uranus (observation 26 on 20111017), which is a
nearly point-like source for \scuba\ that is commonly used as a
primary flux calibrator. The CV daisy pattern was used, with a scan
speed of 155 arcsec\,sec$^{-1}$. We perform several different
reductions of the data to illustrate the purpose of various model
components and the convergence properties of the solution
(Fig.~\ref{fig:pointmaps}). In all cases the maps are produced on a
grid of azimuth (horizontal) and elevation (vertical) offsets from the
position of Uranus (the origin), using 2\,arcsec pixels.

The first, simplest reduction of the data uses only the \model{COM}
model to estimate and remove the common-mode signal in order to
suppress low-frequency noise in the data. After \model{COM}, the
extinction correction is applied (\model{EXT}), and an initial map is
estimated using equal weighting for all of the detectors. This
estimate of \model{AST} is then removed from the data, and the noise
is measured in the residuals to estimate weights for the subsequent
and final iteration. The resulting map after these two iterations is
shown in Fig.~\ref{fig:pointmaps}a. While the peak \snr\ of Uranus is
clearly large ($\sim$160), enabling us to see the faint sidelobes
(circle and cross pattern), the map also has obvious circular streaks
and other large-scale ripples. The circular streaks are due to the
fact that \model{COM} does not account for all of the low-frequency
(i.e., uncorrelated) noise (see Fig.~\ref{fig:pspec}), and therefore
each bolometer leaves a trace of the circular scan pattern in the map,
as their baselines slowly drift independently. A significant
contribution to the larger-scale ripples in the map, however, can be
made by degeneracies in the map solution, as we discuss below.

To illustrate how large-scale ripples can form (and grow), the same
map solution is run for 100 iterations and shown in
Fig.~\ref{fig:pointmaps}b, now exhibiting a strong vertical
gradient. The degeneracy is easy to understand if the time-domain
behaviour of each model component is considered. The top panel of
Fig.~\ref{fig:degeneracy} shows the residual signals for a single
bolometer after 2 (black) and 100 (grey) iterations, which are nearly
identical, yet the change in the estimated
\model{COM}\footnote{\model{COM} has been multiplied by the
time-varying extinction correction to enable direct comparison with
\model{AST}.} (green) and \model{AST} (red) signals between 2 and 100
iterations are large However, it is also clear that the estimated
\model{COM} and \model{AST} signals are complimentary. In other words,
the large change in the \model{AST} signal is cancelled by freedom in
the \model{COM} signal to grow with opposite sign. For comparison, the
bottom panel of Fig.~\ref{fig:degeneracy} shows the telescope pointing
for this section of data, and the shapes of the \model{AST} and
\model{COM} signals match the elevation component, which is aligned
with the gradient in Fig.~\ref{fig:pointmaps}b. Generically, the
calculation of \model{COM} will remove any information on angular
scales that are larger than the array footprint (outline shown in
Fig.~\ref{fig:pointmaps}b for reference), meaning that the map
solution is unconstrained on such large scales.

Attacking the problem of streaks, a simple method of removing residual
(un-correlated) sources of low-frequency noise is to apply a high-pass
filter after the common-mode removal. A third reduction of the data
uses the baseline map-making parameters, as described in
Section~\ref{sec:algorithm}, which adds the \model{FLT} model to
accomplish this task immediately prior to map estimation. We set the
filter edge based on an angular scale of 200\,arcsec, which, given the
scan speed of 155\,arcsec\,sec$^{-1}$, corresponds to a frequency of
0.78\,Hz (note in Fig.~\ref{fig:pspec} that this frequency is slightly
above the $1/f$ knee at 450\,\micron\ after common-mode removal). Now
using the automatic map-based convergence test
(Section~\ref{sec:converge}), the solution converges after 10
iterations (Fig.~\ref{fig:pointmaps}c). Both the circular streaks and
the large-scale gradient are now removed, but they have been replaced
by an obvious, circularly-symmetric ringing pattern about Uranus. The
reason for this ringing is that the hard-edged high-pass filter in
frequency space is equivalent to a $\sinc$ function-like response in
map space. Since the scan pattern is fairly isotropic (scans at all
position angles), and there is a bright point-like source at the
centre, the result is an azimuthally-symmetric $\sinc$ function-like
pattern in the map. The peak flux has also been attenuated to
0.353\,pW in this reduction, compared with 0.358\,pW in the first two
reductions. The gradient resulting from the degeneracy between the
common-mode and map has been removed because the filter scale of
200\,arcsec (3.3\,arcmin) is smaller than the array field-of-view
($\sim$5\,arcmin). If it were greater, degeneracies would again begin
to appear on those larger scales.

This example illustrates the need for constraints in the map solution
in many situations. For calibrators (and other previously known
bright, compact sources), a good, simple prior is to constrain the map
to a value of zero away from the known locations of emission (part of
the \model{AST} model calculation, Section~\ref{sec:ast}).  In
Fig.~\ref{fig:pointmaps}d, a solution is produced in an identical
manner to Fig.~\ref{fig:pointmaps}c, but now setting the map
explicitly to zero beyond a radius of 60\,arcsec from the location of
Uranus (much larger than the FWHM of the main lobe), for all but the
final iteration. In this case, the map converges after 6 iterations,
and the ringing has been effectively removed. The attenuation from the
previous reduction is now removed, and Uranus again has a peak value
of 0.358\,pW. For reference, the Uranus model distributed with
Starlink predicts a peak brightness of 176\,Jy on this date, yielding
an FCF of 492 Jy\,pW$^{-1}$, which is well within the typical range
for \scuba\ of $(492 \pm 67)$\,Jy\,pW$^{-1}$ \citep[see][for
details]{dempsey2012}. Since the map is now flat away from the source,
and constrained to a value of zero, it is appropriate for performing
aperture photometry directly, with no need for an additional reference
annulus.

The way this prior works can be understood from the point of view of
differential measurements. Bolometer data contain information up to
scales corresponding to the filter edge (or the scale of the array
footprint, whichever is smallest). In this example, the relevant scale
is 200\,arcsec. Since the map is constrained to zero within 60\,arcsec
of the peak of Uranus (well within 200\,arcsec), the solution is able
to accurately reconstruct the differential peak intensity of Uranus
with respect to this constrained background. This approach to
map-making is similar to that employed for poorly cross-linked scans
of compact (though resolved) sources by \citet{wiebe2009} using BLAST
data.
